---
title: "R Notebook"
output: html_notebook
---

# metapopultion daynmic

## gettin the data

```{r, setup, include=FALSE}

knitr::opts_knit$set(root.dir = '/Users/yuval/Downloads')
```

```{r}
data <- read.csv('write_up_1_data.csv')
library(ggplot2)
```

## immigration and extinction

for each island I want to calculate it immigration rate and extinction rate for that I will calclate the avarge length of time from extinction to imigration (imigration rate) and from imigration to extinction (extinction rate)

```{r}
# Load necessary library
library(dplyr)

# Convert columns x1 to x30 to binary data
data <- data %>%
  mutate(across(X1:X30, ~ ifelse(. > 0, 1, 0)))

# Function to calculate the mean length of continuous sets of 1s
mean_length_continuous_ones <- function(row ,value) {
  values <- as.numeric(row)
  rle_values <- rle(values)
  lengths_ones <- rle_values$lengths[rle_values$values == value]
  if (length(lengths_ones) > 0) {
    return(mean(lengths_ones))
  } else {
    return(NA)
  }
}

# Apply the function to each row and add the result as a new column
data1 <- data %>%
  rowwise() %>% 
  mutate(extinction_rate = 1/ mean_length_continuous_ones(c_across(X1:X30) ,1) , colnilzation_rate =  1/ mean_length_continuous_ones(c_across(X1:X30) ,0)) %>%
  ungroup()

# Display the modified data fram
meta_comunity <-  select(data1 ,island, size, extinction_rate, colnilzation_rate)
meta_comunity
```

## connction between island size and birth rate

```{r}
model_birth <- glm(birth.rate ~ size, data = data, family = Gamma(link = "log"))
summary(model_birth)
```

```{r}
model_death  <- glm(death.rate ~ size, data = data, family = Gamma(link = "log"))
summary(model_death)
```

```{r}
ggplot(data, aes(x = size)) +
  geom_point(aes(y = birth.rate), color = "blue") +
  geom_point(aes(y = death.rate), color = "red") +
  geom_smooth(aes(y = birth.rate), method = "lm", color = "blue") +
  geom_smooth(aes(y = death.rate), method = "lm", color = "red") +
  labs(x = "Size", y = "Rate", title = "Effect of Size on Extinction and Colonization Rates")
```

## connection between size and extinction rate

```{r}
library(ggplot2)
png(file="size_effect.png",width=600, height=350)
ggplot(meta_comunity, aes(x = size)) +
  geom_point(aes(y = extinction_rate), color = "blue") +
  geom_point(aes(y = colnilzation_rate), color = "red") +
  geom_smooth(aes(y = extinction_rate), method = "lm", color = "blue") +
  geom_smooth(aes(y = colnilzation_rate), method = "lm", color = "red") +
  labs(x = "Size", y = "Rate", title = "Effect of Size on Extinction and Colonization Rates")
dev.off()
```

```{r}

```

now to explore the connection between colonilizatoin / extinction probabitly and patch_ocipancy

```{r}
# Load necessary library
library(dplyr)

# Function to create new columns D1 to D29 based on the given conditions

create_new_columns <- function(df) {
  for (i in 2:30) {
    df <- df %>%
      mutate(!!paste0("X", i-1) := case_when(
        get(paste0("X", i)) == 0 & get(paste0("X", i-1)) == 1 ~ "ext",
        get(paste0("X", i)) == 1 & get(paste0("X", i-1)) == 0 ~ "col",
        get(paste0("X", i)) == 0 & get(paste0("X", i-1)) == 0 ~ "non",
        get(paste0("X", i)) == 1 & get(paste0("X", i-1)) == 1 ~ "stay",
        TRUE ~ NA_character_
      ))
  }
  return(df)
}
# Apply the function to the df data frame
df_new <- create_new_columns(data)  %>% 
  select(island, size, birth.rate, death.rate, starts_with("X"))

# Display the new data frame
print(df_new)
write.csv(df_new, "newdata.csv", row.names=FALSE)

```

```{r}
library(tidyr)
library(dplyr)

# Assuming df is already loaded in your environment
df_long <- df_new  %>%
  select(-birth.rate, -death.rate ,-X30) %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "category",
               values_to = "status")

# Display the first few rows of the transformed data frame
head(df_long)
```

```{r}
df_sum <- data %>%
  select(starts_with("X")) %>%
  summarise(across(everything(), ~ sum(. == 1, na.rm = TRUE))) %>%
  pivot_longer(cols = everything(),
               names_to = "category",
               values_to = "path_ocopncy") %>% filter(category != "X30") %>% mutate(path_ocopncy = path_ocopncy /12)

# Display the new data frame
full <- merge(df_sum ,  df_long , by ="category" )
```

```{r}
colnization_data <- full %>% filter(status == "non" | status == "col") %>% mutate(status_index = case_when(status == "col" ~1 , TRUE~0))
conization_model <- glm(status_index ~ path_ocopncy + size, 
                        data = colnization_data, 
                        family = binomial(link = "logit"))

# Display the summary of the model
summary(conization_model)
```

```{r}
extinction_data <- full %>% filter(status == "stay" | status == "ext") %>% mutate(status_index = case_when(status == "ext" ~1 , TRUE~0))
extinction_model <- glm(status_index ~ path_ocopncy + size, 
                        data = extinction_data, 
                        family = binomial(link = "logit"))

# Display the summary of the model
summary(extinction_model)
```

```{r}
meta_comunity
```

```{r warning=FALSE}
create_whgith_list <- function(Simulation){
  whight <- c()
  for (i in 1:nrow(Simulation)) {
    if (Simulation$current[i] == 1) {
      whight <- c(whight, Simulation$extinction_rate[i])
  } else {
      whight <- c(whight, Simulation$colonization_rate[i])
  }
  }
  return(whight)

}

step <- function(Simulation ,time){
  weights = create_whgith_list(Simulation)
  d_time <- rexp(1, rate = sum(weights))
  random_number <- sample(1:length(weights), 1, prob=weights)[1]
  Simulation$current[random_number] <- 1- Simulation$current[random_number]
  
  return(list(Simulation = Simulation , time = time + d_time) )
}
record <- data.frame(time = 0, sum = 8)
Simulation <- mutate(meta_comunity , current = 1 ) 
time = 0
for (i in 1:1000) {
  sim <-step(Simulation ,time)
  Simulation = sim$Simulation
  time = sim$time
  new_row <- data.frame(time = time, sum = sum(Simulation$current))
  record <- rbind(record, new_row)
  if (sum(Simulation$current) <1 |  time > 300){
    print("end")
    break
  }
}
stable_record <- filter(record , time >100)
png(file="Simulation.png",width=600, height=350)

 ggplot(record, aes(x = time, y = sum/12)) +
  geom_line(color = "red") +
  labs(title = "Change of island ocupncy Over Time",
       x = "Time",
       y = "patch ocupncy") +
  theme_minimal()+ geom_hline(yintercept=0 ) +geom_hline(yintercept= mean(stable_record$sum/12),color = "blue")
 dev.off()
```

```{r}
mean(meta_comunity$extinction_rate)
mean(meta_comunity$colnilzation_rate)

mean(meta_comunity$colnilzation_rate) /mean(meta_comunity$extinction_rate) 
```

```{r}
library("dplyr") 

data2 <- data %>% select(-island , -size , -birth.rate ,-death.rate) 
  sum(data2)/(30*12)

```

```{r}
# Summing columns x1 to x30 into a new data frame
# Summing each column x1 to x30 separately
sum_each_column <- colSums(data[, paste0("X", 1:30)])
sum_df <- data.frame(index = 1:length(sum_each_column), column = names(sum_each_column), sum = sum_each_column)

# Display the sums of each column
 ggplot(sum_df, aes(x = index, y = sum )) +
  geom_line() +
  labs(title = "Change of island ocupncy Over Time",
       x = "Time",
       y = "patch ocupncy") +
  theme_minimal()+ geom_hline(yintercept=0) +geom_hline(yintercept= mean(sum_df$sum))
```

```{r}
data
```
